Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2015-09-20 pkg/glmertree/paper/report.Rnw
added working paper version (report.Rnw, ref.bib) using JSS style
small touch-ups in application R code, joint=FALSE should not be necessary
in the current glmertree version



git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@796 255f4249-2263-4417-b1ea-8498779e27a7

\documentclass[nobf,doc]{apa}
\usepackage{amssymb}
\usepackage{amsmath,amsthm,tabularx,curves,texdraw,psfrag}
\usepackage{apacite}
\usepackage[english]{babel}
\usepackage{lscape,dcolumn,hhline}
\usepackage{graphicx,epic,eepic,rotating}
\usepackage[normalem]{ulem}
\usepackage{fancyhdr}
\usepackage{eurosym,bbding}
\usepackage{verbatim}
\usepackage{ctable}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{lineno}
\linespread{1.2}
\usepackage{tabto}

@@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}
\leftheader{Treatment-Subgroup Interactions with GLMM Trees}
\rightheader{Treatment-Subgroup Interactions with GLMM Trees}
\author{M. Fokkema$^1$, N. Smits$^2$, A. Zeileis$^3$, T. Hothorn$^4$, H. Kelderman$^5$}
\affiliation{$^1$Universiteit Leiden, $^2$Universiteit van Amsterdam, $^3$Universit\"{a}t Innsbruck, $^4$Universit\"{a}t Z\"{u}rich, $^5$Universiteit Leiden and Vrije Universiteit, Amsterdam}

\acknowledgements{The authors would like to thank Prof.~Pim Cuijpers, Prof.~Jeanne Miranda, Dr.~Boadie Dunlop, Prof.~Rob DeRubeis, Prof.~Zindel Segal, Dr.~Sona Dimidjian, Prof.~Steve Hollon and Erica Weitz for granting access to the dataset for the application. The work for this paper was partially done while MF, AZ and TH were visiting the Institute for Mathematical Sciences, National University of Singapore in 2014. The visit was supported by the Institute.}

@@

\maketitle
\pagewiselinenumbers

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation. GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

\section{Introduction}

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \cite<e.g., >{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citeA{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citeNP{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citeNP{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \cite{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM). GLM-based MOB has been successfully applied by \citeA{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies. Also, \citeA{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.

In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure. However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \cite<e.g., >{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \cite<e.g., >{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \cite<e.g., >{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \cite<which is not done in other tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.

\subsection{Artificial motivating dataset}

@@

\section{General modeling framework}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \cite{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citeA{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

To find partitions and better-fitting local GLMs, the MOB algorithm tests for parameter instability. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find partitions, the MOB algorithm cycles iteratively through the following steps \cite{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. By definition, the empirical scores of all observations in a dataset sum to zero, and when the model is correctly specified, the expected value of the score for each observation is also zero. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \cite<c.f., >{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \cite{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \cite{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citeA{BrykyRaud92}, for example.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

\begin{figure}[t!]
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\section{Empirical evaluation}

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citeA{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. That is, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[t!]
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

\section{Application to patient-level meta-analysis dataset on the effects of treatments for depression}

@@

<<depression-trees>>=
hrs <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))

>>>>

\documentclass[nojss]{jss}
\usepackage{amssymb,amsmath,amsthm}

\usepackage{thumbpdf}
\shortcites{CuijyWeit14,DrieySmit14}

@@

\title{Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees}
\Shorttitle{Detecting Treatment-Subgroup Interactions in Clustered Data with GLMM Trees}

\author{Marjolein Fokkema\\Universiteit Leiden \And
        Niels Smits\\Universiteit van Amsterdam \And
	Achim Zeileis\\Universit\"{a}t Innsbruck \AND
	Torsten Hothorn\\Universit\"{a}t Z\"{u}rich \And
	Henk Kelderman\\Universiteit Leiden\\Vrije Universiteit, Amsterdam}
\Plainauthor{Marjolein Fokkema, Niels Smits, Achim Zeileis, Torsten Hothorn, Henk Kelderman}

\Abstract{
  Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Several tree-based algorithms have been developed for the detection of such treatment-subgroup interactions. In many instances, however, datasets may have a clustered structure, where observations are clustered within, for example, research centers, studies or persons. In the current paper we propose a new algorithm, generalized linear mixed-effects model (GLMM) trees, that allows for detection of treatment-subgroup interactions, as well as estimation of cluster-specific random effects. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM tree and compare it with that of MOB without random-effects estimation. GLMM tree was found to have a much lower Type I error rate than MOB trees without random effects (4\% and 33\%, respectively). Furthermore, in datasets with treatment-subgroup interactions, GLMM tree recovered the true treatment subgroups much more often than MOB without random effects (in 90\% and 61\% of the datasets, respectively). Also, GLMM tree predicted treatment outcome differences more accurately than MOB without random effects (average predictive accuracy of .94 and .88, respectively). We illustrate the application of GLMM tree on a patient-level dataset of a meta-analysis on the effects of psycho- and pharmacotherapy for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
}
\Keywords{model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees}

\Address{
  Marjolein Fokkema\\
  Universiteit Leiden\\
  Faculteit der Sociale Wetenschappen\\
  Instituut Psychologie, Methodologie \& Statistiek\\
  Pieter de la Court gebouw\\
  Wassenaarseweg 52\\
  2333 AK Leiden, The Netherlands\\
  E-mail: \email{m.fokkema@fsw.leidenuniv.nl}\\
  % URL: \url{http://www.socialsciences.leiden.edu/psychology/organisation/staff/fokkemam.html}	
}

@@

\section{Introduction} \label{sec:intro}

In research assessing the efficacy of treatments for somatic and psychological disorders, the one-size-fits-all paradigm is slowly losing ground, and personalized medicine is becoming increasingly important. Stratified medicine presents the challenge of finding which patients respond best to which treatments. This can be referred to as the detection of treatment-subgroup interactions \citep[e.g.,][]{DoovyDuss14}. In most cases, treatment-subgroup interactions are studied using linear models, such as factorial analysis of variance techniques, in which potential moderators have to be specified a-priori, have to be checked one at a time, and continuous moderator variables have to be discretized a-priori. This may hamper identification of which treatments work best for whom, especially when there are no a-priori hypotheses about treatment-subgroup interactions. As noted by \citet{KreayFran06}, there is a need for methods that generate, instead of test, hypotheses and that are specifically directed at the detection of treatment interactions.

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \citep{StroyMall09}. Several promising tree-based algorithms for the detection of treatment-subgroup interactions have been developed (e.g., \citealp{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}; see \citealp{DoovyDuss14} for an overview). Among these methods, model-based recursive partitioning (MOB; \citealp{ZeilyHoth08}) seems to be the most flexible tool for detecting treatment-subgroup interactions, as it offers a generic inferential framework that can be coupled with a broad range of parametric modeling strategies fitted by M-type estimators \citep{ZeilyHoth08}. Specifically, this model class encompasses the generalized linear model (GLM). GLM-based MOB has been successfully applied by \citet{DrieySmit14} in the detection of subgroups of patients with depression, which differ in the effect of two different psychotherapies. Also, \citet{SeibyZeil15} applied MOB for in the detection of subgroups of patients with amyotrophic lateral sclerosis (ALS), which differ in the effect of treatment with Riluzole.

In many cases, researchers may want to detect treatment-subgroup interactions in datasets with a clustered structure. However, none of the aforementioned tree-based algorithms allow for taking into account the clustered structure of a dataset. For example, in individual-level patient data meta-analyses, in which datasets of multiple trials evaluating the effects of the same treatments are pooled \citep[e.g.,][]{KoopyHeij07}. In such analyses, the clustered structure of the dataset should be taken into account by including study-specific effects in the model, prompting the need for modeling random effects \citep[e.g.,][]{CoopyPatt09,HiggyWhit01}. Likewise, longitudinal datasets, and datasets from multi-center trials typically also require modeling of random effects. Ignoring the clustered structure of datasets may lead to biased inference due to underestimated standard errors \citep[e.g.,][]{BrykyRaud92,NooryOpde05}. More specifically, when the interest is in subgroup detection, ignoring random effects on the outcome variable may result in the detection of spurious subgroups \citep[e.g.,][]{SelaySimo12}.

In the current paper, we present a tree-based algorithm for detecting treatment-subgroup interactions, which takes the clustered nature of datasets into account. The algorithm combines MOB with random-effects estimates and therefore accounts for both the clustering structure \citep[which is not done in other tree-based treatment-subgroup interaction detection methods, e.g.,][]{ZeilyHoth08, SuyTsai09, DussyMech14} and the treatment effect estimation in models with continuous and non-continuous response variables \citep[which is not available in previously suggested regression trees with random effects, e.g.,][]{HajjyBell11,SelaySimo12}.

In Section~\ref{sec:framework}, we will introduce the existing frameworks for estimating treatment effects: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Section~\ref{sec:evaluation}, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Section~\ref{sec:application}, we illustrate the application of the algorithm, using an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citet{CuijyWeit14}. But before we discuss the frameworks for estimating treatment effects, we will introduce an artificial motivating data set with which the frameworks and methods to be discussed, will be illustrated.

\subsection{Artificial motivating dataset} \label{sec:data}

@@

\section{General modeling framework} \label{sec:framework}

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.5\textwidth}
<<exampleglm, fig=TRUE, height=5, width=5>>=
plot(exampleglm, tp_args = list(mainlab = "Full sample (N = 150)"))
@
\caption{Example of a normal GLM (with fixed effects only) for treatment outcomes, based on the artificial motivating dataset ($N=150$). The dot for Treatment~1 represents the first, and the slope of the regression line represents the second element of $\beta$.}
\label{fig:fixedeffects}
\end{figure}

@@

The rationale behind MOB is that a global model for all observations, like the GLM in Equation~\ref{eq:expected_value} and~\ref{eq:fixedeffects}, may not describe all data well, and when additional covariates are available it may be possible to partition the dataset with respect to these covariates, and find a better model in each cell of the partition \citep{ZeilyHoth08}. This is reminiscent of the classification and regression tree (CART) algorithm of \citet{BreiyFrie84}, which splits the dataset into subsets, for which the distributions of the outcome variable are most different. However, CART trees detect differences in constant fits across terminal nodes, whereas MOB trees detect differences in parametric models across terminal nodes.

To find partitions and better-fitting local GLMs, the MOB algorithm tests for parameter instability. When the partitioning is based on a GLM, instabilities are differences in $\hat{\beta}$ across partitions of the dataset, which are defined by one or more auxiliary covariates not included in the linear predictor. To find partitions, the MOB algorithm cycles iteratively through the following steps \citep{ZeilyHoth08}: (1) fit the parametric model to the dataset, (2) test for parameter instability over a set of partitioning variables, (3) if there is some overall parameter instability, split the dataset with respect to the variable associated with the highest instability, (4) repeat the procedure in each of the resulting subgroups.

More specifically, in step (2), to test for parameter instability, the so-called \textit{scores} are computed, using the score function. By definition, the empirical scores of all observations in a dataset sum to zero, and when the model is correctly specified, the expected value of the score for each observation is also zero. Under the null hypothesis of parameter stability, the scores do not systematically deviate from the expected value of zero, when the observations are ordered by the values of a potential partitioning variable $U_k$ \citep[c.f.,][]{MerkyZeil13}. To statistically test whether the scores systematically deviate from zero with respect to variable $U_k$, the class of generalized M-fluctuation tests is used \citep{Zeil05,ZeilyHorn07}.

If the null hypothesis of parameter stability in step (2) can be rejected, that is, if at least one of the partitioning variables $U_{k}$ has a p-value for the M-fluctuation test below the pre-specified significance level $\alpha$, the dataset is partitioned into two subsets in step (3). In step (3), a binary partition is created using $U_{k*}$, the variable with the minimal p-value in step (2). The split point for $U_{k*}$ is selected, by taking the value that minimizes the sum of the values of the objective function in both partitions \citep{ZeilyHoth08}. In step (4), steps (1) through (3) are repeated in each partition, until the null hypothesis of parameter stability can no longer be rejected.

@@

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=\textwidth}
<<exampleglmtree, fig=TRUE, width=12, height=9>>=
plot(exampleglmtree)
@
\caption{Example of a tree representation of model-based recursive partition, based on the artificial motivating dataset. Three additional covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables.}
\label{fig:example_mobtree}
\end{figure}

@@

\subsection{GLMM}
When a dataset contains observations from multiple clusters (e.g., trials, research centers, or individuals in longitudinal datasets), the GLM in Equation~\ref{eq:fixedeffects} may be extended to include cluster-specific, or random effects, and the model becomes a GLMM:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
Where $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, with every element being the random intercept for cluster $m$. Within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML), as described in \citet{BrykyRaud92}, for example.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of \citet{HajjyBell11} and \citet{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

\begin{figure}[t!]
\centering
<<exampleglmmtree, fig=TRUE, width=9, height=7>>=
plot(exampleglmmtree)
@
\caption{GLMM tree of the motivating example dataset. Three covariates (anxiety questionnaire score, duration of depressive symptoms at baseline in months and age) were used as potential splitting variables, and the clustering structure was taken into account by estimating random intercepts.}
\label{fig:example_glimmertree}
\end{figure}

@@

\section{Empirical evaluation} \label{sec:evaluation}

@@

\subsubsection{Datasets with treatment-subgroup interactions}
For generating datasets with treatment-subgroup interactions, we used a treatment-subgroup interaction design from \citet{DussyMech14}, which is also depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by their values on the partitioning variables $U_2$, and $U_1$ or $U_5$. That is, $U_1$, $U_2$ and $U_5$ are true partitioning variables, whereas other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

\begin{figure}[t!]
\centering
\setkeys{Gin}{width=0.8\textwidth}
<<dgp-tree, fig=TRUE, height=5, width=7>>=
fig4 <- party(
  partynode(1L,
    split = partysplit(2L, breaks = 30),
    kids = list(
      partynode(2L,
        split = partysplit(1L, breaks = 17),
        kids = list(
          partynode(3L, info = c(
	    expression(''),
	    expression(beta[j0] == '17.5'),
	    expression(''),
	    expression(beta[j1] == -'5.0'),
	    expression(''),
	    expression(d[j] == -'1.0')
	  )),
          partynode(4L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )))),
      partynode(5L,
        split = partysplit(1L, breaks = 63),
        kids = list(
          partynode(6L, info = c(
	    expression(''),
	    expression(beta[j0] == '30.0'),
	    expression(''),
	    expression(beta[j1] == '0.0'),
	    expression(''),
	    expression(d[j] == '0.0')
	  )),
          partynode(7L, info = c(
	    expression(''),
	    expression(beta[j0] == '42.5'),
	    expression(''),
	    expression(beta[j1] == '5.0'),
	    expression(''),
	    expression(d[j] == '1.0')
	  )))))),
  data.frame(U1 = numeric(0), U2 = numeric(0))
)
plot(fig4, tp_args = list(FUN = identity, width = 9), tnex = 1.5)
@
\caption{Data-generating model for treatment-subgroup interactions. Parameter $d$ denotes the standardized mean difference between the outcomes of Treatment~1 and~2 (i.e., $\beta_{j1} / \sigma_{\epsilon}$).}
\label{fig:modelC}
\end{figure}

@@

\begin{enumerate} 
\item Three levels for the total number of observations: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts are drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two different levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes, in subgroups with differential treatment effects. The levels for mean differences in subgroups with differential treatment effect were $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citealp{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$).
\end{enumerate}

@@

\section{Patient-level meta-analysis of depression treatments} \label{sec:application}

@@

<<depression-trees>>=
## offset
metadata$HRSDfit <- fitted(lm(HRSDt1 ~ HRSDt0, data = metadata))
